\section{Conclusions and future work}

Currently, parse results that are not part of the final structure of
the buffer are discarded.  When the user is not using an editor mode
that automatically balances characters such as parentheses and double
quotes, inserting such a character often results in a large number of
parse results being discarded, only to have to be created again soon
afterward, when the user inserts the balancing character of the pair.
We can avoid this situation by keeping parse results that are not part
of the final structure in the cache, in the hopes that they will again
be required later.  We then also need a strategy for removing such
parse results from the cache after some time, so as to avoid that the
cache grows without limits.

Parsing \commonlisp{} source code is only the first step in the
analysis of its structure.  In order to determine the role of each
symbol and other information such as indentation, further analysis is
required.  Such analysis requires a \emph{code walker}, because the
role of a symbol may depend on the definitions of macros to which it
is an argument.  Similarly, computing standard
\emph{indentation}, also requires further analysis.  To implement this
code walker, we consider using the first phase of the \cleavir{}
compiler framework.%
\footnote{\cleavir{} is currently part of \sicl{}.  See the directory
  named Code/Cleavir in the \sicl{} repository.}

We plan to investigate the use of a new implementation of
\emph{first-class global environments}
\cite{Strandh:2015:ELS:Environments}.  This new implementation of the
existing \clos{} protocol would use \emph{incremental differences} to
the \emph{startup environment}%
\footnote{Recall that the startup environment is the environment in
  which the compiler was invoked.}
so as to define a \emph{compilation environment}%
\footnote{Recall that the compilation environment is the environment
  used by the compiler for definitions and side effects of the
  compilation process.}
that is different for each top-level form in the editor buffer.  This
technique would allow us to restart the compiler in an appropriate
environment without having to process the entire buffer from the
beginning.

The combination of the use of the first pass of the \cleavir{}
compiler framework and the use of incremental first-class global
environments will allow us to handle compile-time evaluation of
certain top-level forms in a way that corresponds to the semantics of
the file compiler.  In particular, imperative environment operations
such as changing the current package or modifying the readtable in the
middle of a buffer will have the expected consequences, but only to
subsequent forms in the buffer.

A more precise analysis of \commonlisp{} code opens the
possibility for additional functionality that requires knowledge about
the role of each expression.  In particular, such an analysis could be
the basis for sophisticated code transformations such as variable
renaming and code refactoring.
