\chapter{General \commonlisp{} style guide}

\section{Purpose of style restrictions}

The purpose of imposing a particular style is based on a few simple
facts that hold true for both natural languages and programming
languages:

\begin{itemize}
\item The set of all idiomatic phrases is a tiny subset of the set of
  all grammatical phrases.
\item The main purpose of these phrases is to serve as communication
  between people.
\end{itemize}

To illustrate the first fact, consider a natural language such as
English.  In English, we say ``tooth brush'', but ``dental floss''.
The words ``dental brush'' and ``tooth floss'' would be perfectly
grammatical, but they are not used.  A person trying to communicate
with other people must use the words that have been widely agreed
upon, even though some other words are perfectly legitimate.  It might
seem that such idiosyncrasies would be limited to languages with
multiple heritage such as English, but that is not the case.  In
French, we say ``brosse à dents'', ``pâte dentifrice'', and ``fil
dentaire''.  There are nine reasonable combinations, but only one is
used.

The same thing is true for programming languages.  The community has
collectively decided on a particular subset of all the grammatical
phrases, and a programmer who wishes to communicate with other
programmers should stick to that subset.

It should also be emphasized that the choice of idioms is different in
different languages.  An example from natural languages would be that
in English we say ``I was my hands'', in French ``I wash myself the
hands'', and in Swedish we say ``I wash the hands [on myself]''.  Just
as it would be pointless trying to use an idiom form one language in a
translated version in a different language, it is as pointless to
translate idioms from one programming language to a different
programming language.

Finally, the choice of what phrases are idioms and what phrases are
not is almost totally arbitrary, and based on coincidences of
history.  Therefore it is rarely productive to ask oneself why a
particular phrase is an idiom and a different one is not.  There is no
possible enlightening answer to such a question.


\section{Width of a line of code}

Horizontal space is a precious resource that should not be wasted.
The width of a line should preferably not exceed 80 characters.  This
limit used to be hard, because some printers or printer drivers would
truncate longer lines.  Since it is less common to print code these
days, the limit is now soft.  The purpose of keeping lines somewhat
short is so that it is possible on a reasonable monitor to display two
documents side by side.  One document is typically a \commonlisp{}
source file, and the other document is typically the buffer containing
interactions with the \commonlisp{} system.

The systematic use of long lines makes the practice of displaying two
documents side by side impossible, or at least impractical.  If a
single monitor is used, the programmer then has to flip back and forth
between the source code and the interaction loop.  When two monitors
are used, the effect is to waste half a monitor that could otherwise
be used for displaying documentation or something else.

\section{Blank lines}

A single blank line is common in the following situations:

\begin{itemize}
\item Between two top-level forms.
\item Between a file-specific comment and the following top-level
  form.
\item Between a comment for several top-level forms and the first
  of those top-level forms.
\end{itemize}

A single blank line \emph{may} occur inside a top-level form to
indicate the separation of two blocks of code concerned with different
subjects, but it would be more common to put those two blocks of code
in separate functions.

There should never be any instance of two consecutive blank lines, and
the last line of the file should not be blank.

\section{\texttt{car}, \texttt{cdr}, \texttt{first}, etc are for
  \texttt{cons} cells}

The \commonlisp{} standard specifies that the function \texttt{car},
\texttt{cdr}, \texttt{first}, \texttt{second}, \texttt{rest}, etc
return \texttt{nil} when \texttt{nil} is passed as an argument.  This
fact should mostly be considered as a historical artifact and should
not be systematically exploited.  Take for instance the following
code:

\begin{verbatim}
(if (first x) ...)
\end{verbatim}

To the compiler, it means ``execute the false branch of the \texttt{if}
when either \texttt{x} is \texttt{nil}, or when \texttt{x} is a list
whose first element is \texttt{nil}''.

To the person reading the code, it means something different
altogether, namely ``\texttt{x} holds a non-empty list of Boolean
values, and the false branch of the \texttt{if} should be executed
when the first element of that list is
\emph{false}. See also \refSec{sec-coding-style-meanings-of-nil}.

\section{Different meanings of \texttt{nil}}
\label{sec-coding-style-meanings-of-nil}

Consider the following local variable bindings:

\begin{verbatim}
(let ((x '())
      (y nil)
      z)
  ...)
\end{verbatim}

To the compiler, the three are equivalent.  To a person reading the
code, they mean different things, however:

\begin{itemize}
\item The initialization of \texttt{x} means that \texttt{x} holds a
  \emph{list} that is initially empty.
\item The initialization of \texttt{y} means that \texttt{y} holds a
  Boolean value or a default value that may or may not change in the
  body of the \texttt{let} form.
\item The absence of initialization of \texttt{y} means that no
  initial value is given to \texttt{z}.  In the body of the
  \texttt{let} form, the variable \texttt{z} will be assigned to
  before it is used.
\end{itemize}

The following body of the \texttt{let} form corresponds to the
expectations of the person reading the code:

\begin{verbatim}
(let ((x '())
      (y nil)
      z)
  ...
  (push (f y) x)
  ...
  (unless y (setf y (g x)))
  ...
  (setf z (h x))
  ...)
\end{verbatim}

The following body of the \texttt{let} form violates the expectations
of the person reading the code:

\begin{verbatim}
(let ((x '())
      (y nil)
      z)
  ...
  (push (f y) z)     ; z is used before it is assigned.
  ...
  (unless x          ; x is treated as a Boolean.
    (setf y (g x)))
  ...
  (push (f x) y)     ; y is treated as a list.
  ...)
\end{verbatim}

\section{Tests in conditional expressions}

The \emph{test} of a conditional expression should be a (possibly
generalized) Boolean expression.  The following expressions correspond
to the expectations of the person reading the code:

\begin{verbatim}
(if visited-p ...)
(when (member ...) ...)
(cond ((plusp x) ...) ...)
\end{verbatim}

The following code violates the expectation:

\begin{verbatim}
(let ((item (find ...)))
  (when item ...))
\end{verbatim}

because \texttt{item} is not a (generalized) Boolean value.  It is an
item returned by \texttt{find}, though there is an \emph{out of band}
value (\texttt{nil}) indicating that no item was found by
\texttt{find}.  In this case, the corresponding code that corresponds
to the expectations would look like this:

\begin{verbatim}
(let ((item (find ...)))
  (unless (null item) ...))
\end{verbatim}

\section{General structure of recursive functions}

When possible, a recursive function should be structured like a
mathematical proof by induction.  By that we mean that the special
case should be handled \emph{first} so as to reassure the person
reading the code that this case can be handled correctly by the
function.


So for instance, assume we have want to write a function that counts
the number of atoms in a tree, we should not write it like this:

\begin{verbatim}
(defun count-atoms (tree)
  (if (consp tree)
      (+ (count-atoms (car tree))
         (count-atoms (cdr tree)))
      1))
\end{verbatim}

but rather

\begin{verbatim}
(defun count-atoms (tree)
  (if (atom tree)
      1
      (+ (count-atoms (car tree))
         (count-atoms (cdr tree)))))
\end{verbatim}

Even when the base case does not return anything useful, it should be
handled first.  The following code violates the expectations:

\begin{verbatim}
(defun map-conses (function tree)
 (unless (atom node)
   (funcall function node)
   (traverse (car node))
   (traverse (cdr node))))
\end{verbatim}

and should be written like this instead:

\begin{verbatim}
(defun map-conses (function tree)
  (if (atom node)
      nil ; nothing to do
      (progn (funcall function node)
             (traverse (car node))
             (traverse (cdr node)))))
\end{verbatim}

though, admittedly, this example is a little too simple to illustrate
the importance of this rule.

\section{Using \texttt{car} and \texttt{cdr} vs. using \texttt{first}
  and \texttt{rest}}

While the two functions \texttt{car} and \texttt{first} have the exact
same definitions, as do \texttt{cdr} and \texttt{rest}, they send very
different messages to the person reading the code.

The functions \texttt{car}, \texttt{cdr}, etc., should be avoided when
the argument is to be considered as a \emph{list}, and should be
reserved for other uses of \texttt{cons} cells such as for
\emph{trees} or \emph{pairs} of values.

It follows that the two families of functions should never be mixed
for the same argument.

\section{Commenting}

Comments are meant to be read by the maintainers of the code.  One can
therefore safely assume that the reader is quite familiar with the
\commonlisp{} language and the main structure of \clim{} and \sysname{}.

Comments should be used to explain aspects of the code that are not
obvious from reading the code itself.  A comment is the ideal place to
introduce definitions of concepts that must be understood in order for
the code to make sense.  If the code is structured in a particular way
for performance reasons, then a comment is a good place to indicate
such a fact, so as to avoid that the maintainer be tempted to
introduce modifications that alter this structure.

\section{Designators for symbol names}

Always use uninterned symbols (such as \texttt{\#:hello}) whenever a
string designator for a symbol name is called for.  In particular,
this is useful in \texttt{defpackage} and \texttt{in-package} forms.

Using the upper-case equivalent string makes the code break whenever
the reader is case-sensitive (and it looks strange that the designator
has a different case from the way symbol that it designates is then
used), and using keywords unnecessarily clutters the keyword package.

\section{Docstrings}

We believe that it is a bad idea for an implementation of a Lisp
system to have docstrings in the same place as the definition of the
language item that is documented, for several reasons.  First, to the
person reading the code, the docstring is most often noise, because it
is known from the standard what the language item is about.  Second,
it often looks ugly with multiple lines starting in column 1 of the
source file, and this fact often discourages the programmer from
providing good docstring.  Third, it makes internationalization
harder.

For this reason, we will provide language-specific files containing
all docstrings of Common Lisp in the form of calls to \texttt{(setf
documentation)}.

\section{Naming and use of slots}

In order to make the code as safe as possible, we typically do not
want to export the name of a slot, whereas frequently, the reader or
the accessor of that slot should be exported.  This restriction
implies that a slot and its corresponding reader or accessor cannot
have the same name.  Several solutions exist to this problem.  The one
we are using for \sysname{} is to have slot names start with the
percent character (`\texttt{\%}').  Traditionally, a percent character
has been used to indicate some kind of danger, i.e. that the
programmer should be very careful before directly using such a name.
Client code that attempts to use such a slot would have to write
\texttt{package::\%name} which contains two indicators of danger,
namely the double colon package marker and the percent character.

Code should refer to slot names directly as little as possible.  Even
code that is private to a package should use an internal protocol in
the form of readers and accessors, and such protocols should be
documented and exported whenever reasonable.  It sometimes good
practice to have multiple accessors for a slot, one for internal
purposes and one for use by client code.  This practice allows for
\code{:before}, \code{:after}, and \code{:around} methods on one
accessor but not the other.

\section{Using other packages}

The \code{:use} option of \code{defpackage} and the \code{use-package}
function should be restricted to the \code{common-lisp} package as
much as possible.  The reason for this restriction is that using a
package this way represents a commitment to accepting all exported
symbols of that package, current and future, whereas in most cases
there is no guarantee that future modifications of the package will
not introduce symbol conflicts.  If it is desired to avoid explicit
package prefixes in some cases, then it is better to use the
\code{:import-from} option of \code{defpackage} to import an
explicitly-supplied list of symbols.

\section{Conditions, restarts, and reporting}

Conditions should not be simple conditions, because we want condition
reporting to be subject to internationalization.   For the same
reason, the condition reporter should not be part of the
\code{define-condition} form, and instead be written separately in a
file that contains language-specific condition reporters.

Restarts should be provided whenever practical.

\section{Internationalization}

We would like for {\sysname} to have the ability to report messages in
the local language if desired.
For this, we use the library named \texttt{acclimation}.%
\footnote{See https://github.com/robert-strandh/Acclimation}

\section{Threading and thread safety}

Consider the use of locks to be free.  A technique called
``speculative lock elision'' is already available in some processors,
and we predict it will soon be available in all main processors.
